---
title: "Calculating the Area of a Circle Using Numerical Integration in R"
lang: en
categories: [R, Numerical Integration]
date: "2023-09-10"
---

Note: This article is translated from [my Japanese article](https://uchidamizuki.quarto.pub/blog/posts/2023/09/circle-area-by-numerical-integration-in-r.html).

## Introduction

I recently learned that the `integrate()` function in R's stats package[^1] can perform numerical integration of one-dimensional functions. So in this article, referring to [this article](https://mathwords.net/enmensekibun), I tried calculating the area of a circle using integration.

[^1]: The stats package is one of the packages that comes with R by default.

## A function to draw a semicircle

A function that draws a semicircle of radius $r$ can be expressed as @eq-semicircle. Written in R, this becomes the `semicircle()` function.

$$
y = \sqrt{r^2 - x^2} \qquad (-r \le x \le r)
$$ {#eq-semicircle}

```{r}
semicircle <- function(x, radius) {
  sqrt(radius^2 - x^2)
}
```

Let's use the `semicircle()` function we implemented to draw a semicircle with radius $r = 5$.

```{r}
radius <- 5
curve(semicircle(x, radius), -radius, radius,
      asp = 1) # Set the aspect ratio to 1:1
```

## Calculating the area of a circle with numerical integration

Next, let's use the `stats::integrate()` function to numerically integrate the `semicircle()` function we just wrote. For radius $r = 5$, we can confirm that the integral is approximately equal to the area of a semicircle, $\frac{1}{2}\pi r^2$.

```{r}
# Set the radius to 5
radius <- 5
integrate(semicircle, -radius, radius, # Integrate over the range from -radius to radius
          radius = radius)

pi * 5^2 / 2
```

Therefore, we can write `circle_area_approx()`, which approximately computes the area of a circle, as follows (the result of the numerical integration is stored in the `value` element of the return value of `stats::integrate()`).

```{r}
circle_area_approx <- function(radius) {
  out <- integrate(semicircle, -radius, radius,
                   radius = radius)
  out$value * 2
}

circle_area_approx(5)
```

Finally, let's compare the approximate values from `circle_area_approx()` with the theoretical values from `circle_area_true()` ($\pi r^2$) over the range $1 \le r \le 10$.

The calculation confirms that the approximate and theoretical circle areas are nearly equal.

```{r}
circle_area_true <- function(radius) {
  pi * radius ^ 2
}

data <- data.frame(radius = 1:10)
data$circle_area_approx <- sapply(data$radius, circle_area_approx)
data$circle_area_true <- sapply(data$radius, circle_area_true)
knitr::kable(data)
```

## Conclusion

We have seen that the `stats::integrate()` function makes it easy to perform numerical integration of one-dimensional functions.

### A note of caution

The `stats::integrate()` function can also perform numerical integration over the range $-\infty \le x \le \infty$, but be aware that it may not integrate correctly when given large values such as $10^{-6} \le x \le 10^6$.

```{r}
# OK
integrate(dnorm, -Inf, Inf)

# Not OK
integrate(dnorm, -1e6, 1e6)
```
